Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-05 paper.tex
Add in application refs

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024}.
Additionally, there is a rich suite of tools growing around
the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
adrion_community_2020,terasaki_geonomics_2021,
baumdicker_efficient_2021,
korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
wohns_unified_2022,mahmoudi_bayesian_2022,
rasmussen_espalier_2022,
zhang_biobank-scale_2023,zhan_towards_2023} and analysis
packages
% etc
\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
Infererence methods now tend to fall into two classes: those that
sample from a statistical
% TODO add SINGER
model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022}
or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},
and those based on
heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
zhang_biobank-scale_2023}.
Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

>>>>

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Applications using inferred ARGs as input have begun to
appear~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}.
Additionally, there is a rich suite of tools growing around
the \texttt{tskit} ARG library~\citep{ralph_efficiently_2020}
including simulation~\citep{kelleher_efficient_2016,kelleher_efficient_2018,
adrion_community_2020,terasaki_geonomics_2021,
baumdicker_efficient_2021,
korfmann_weak_2022, petr_slendr_2023,tagami_tstrait_2024},
ARG inference~\citep{kelleher_inferring_2019,speidel_method_2019,
wohns_unified_2022,mahmoudi_bayesian_2022,
rasmussen_espalier_2022,
zhang_biobank-scale_2023,zhan_towards_2023} and analysis
packages
% etc
\citep{nowbandegani_extremely_2023,fan_likelihood_2023}.
Infererence methods now tend to fall into two classes: those that
sample from a statistical
model~\citep{rasmussen_genome-wide_2014,hubisz_mapping_2020,mahmoudi_bayesian_2022,
deng_robust_2024}
or parsimony~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022},
and those based on
heuristics~\citep{kelleher_inferring_2019,speidel_method_2019,
zhang_biobank-scale_2023}.
Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.
